Revealing the movement dynamics of hippocampal replay using state space modeling

### 2020-09-29 Article.tex

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
Representations of past and possible future experiences play a critical role in memory and decision-making processes. The hippocampus is known to express such representations during sharp-wave ripple (SWR) events, when ensembles of hippocampal place cells can be activated in time-compressed sequences consistent with prior experiences. Efforts to understand these "replay" events have typically imposed stringent criteria on event significance, and thus analyze only a small subset of events.  As a result, we lack a clear picture of both the prevalence and the range of dynamics associated with replay. Here, we develop a flexible state space model that uses a combination of discrete and latent states to capture the spatial content and time evolution of replay during SWR events. Application of this model reveals that the large majority of SWR events contain coherent spatial content that evolves in accordance with a surprisingly wide range of mental movement speeds.

@@

We note here that the choice of any particular threshold for classifying a SWR is arbitrary, and that the power of our approach lies in part in the ability to assign a confidence value for each dynamic or combinations of dynamics to each moment in time. Our goal in choosing 0.8 was to use a threshold that corresponds to highly probable and to explicitly avoid the arbitrary choice of 0.95, as the 0.95 threshold risks giving the false conclusion that only events that are statistically significant at the 0.05 level are meaningful in the brain. Nonetheless, we also verified that our central results hold with a higher threshold of 0.95.

>>>>

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
Representations of past and possible future experiences play a critical role in memory and decision-making processes. The hippocampus is known to express such representations during sharp-wave ripple (SWR) events, when ensembles of hippocampal place cells can be activated in time-compressed sequences consistent with prior experiences. Efforts to understand these "replay" events have typically imposed stringent criteria on event significance, and thus analyze only a small subset of events (~10-45\%). As a result, we lack a clear picture of both the prevalence and the range of representational dynamics associated with replay. Here, we develop a state space model that uses a combination of movement dynamics to capture the spatial content and time evolution of replay during SWR events. Application of this model reveals that the large majority of SWR events contain coherent spatial content that evolves in accordance with a surprisingly wide range of mental movement speeds.

@@

We note here that the choice of any particular threshold for classifying a SWR is arbitrary, and that the power of our approach lies in part in the ability to assign a confidence value for each dynamic or combinations of dynamics to each moment in time. Our goal in choosing 0.8 was to use a threshold that corresponds to highly probable and to explicitly avoid the arbitrary choice of 0.95, as the 0.95 threshold often leads people to conclude that only events that are statistically significant at the 0.05 level are meaningful in the brain. Nonetheless, we also verified that our central results hold with a higher threshold of 0.95.
